---
title: |-
  Exploratory Data Analysis of
   North Carolina MRIP Datasets
   for 2013
author: "Scott Smith"
date: "Monday, February 23, 2015"
output: pdf_document
---


#Synopsis  
This study aims to examine the NOAA MRIP dataset from 2013 using exploratory data analysis.   
Specific goals include:  

1. Dividing the data into regions, North, South, and Central NC. Based on internal NCDMF divisions.  
2. Analyze variation between regions for both angler specific and region specific trends.  
3. Determine areas for improvement based on comparisons between regions.  


#Data Processing  
A majority of the data processing code has been left out of this document for readability. A document with full code is available on the GitHub repository listed below. The individual *.R files are also provided in the repository for reproducibility.  

[www.GitHub.com/sas6523/NC_MRIP_ANALYSIS](www.GitHub.com/sas6523/NC_MRIP_ANALYSIS)  

## Loading Data  
Data is loaded into a temporary directory using download.file() and unzip().    
```{r, echo=TRUE, warning=FALSE,error=FALSE, message=FALSE}
library(plyr)
##########################################################
# Enter the year for which you wish to run the analysis. 
# At the time of writing, valid dates: 2004-2013
# Previous years used diferent MRFSS Datasets, 
# which will not work here.
year <- "2012"
##########################################################

# This is used as a marker for the date the data were downloaded
today = as.Date(Sys.time())
#today = gsub('-', '', today)

# Form the url from the year variabl and the url template
url <- paste0("http://www.st.nmfs.noaa.gov/st1/recreational/MRIP_Survey_Data/CSV/PS_", 
              year, "_CSV.zip")

# create a temporary directory
td = tempdir()
setwd(td)
if(!exists(paste0("PS_", year, "_CSV.zip"))){
        # create the placeholder file
        tf = tempfile(tmpdir=td, fileext=".zip")
        # download into the placeholder file
        download.file(url, tf)
        # Unzip the file into the temp directory
        unzip(tf, exdir = td)
}

# Create a list of the filenames that were unzipped
files <- list.files(td, pattern = ".csv")
# Create a list of the names without .csv. This will correspond to our df names
names <- gsub(pattern='\\.csv$', '', files, ignore.case=T)

# Load the files we just unziped into dataframes in R
for( i in 1:length(names)){
        a <- paste(names[i], ' <- read.csv( file=\'', files[i], 
                   '\', header=TRUE, sep=\',\')', sep='')
        # print( a )
        eval(parse( text=a ))   
}
```

At this point, we need to verify that the files were read correctly:  

* Files List:  
** `r files`  
* Data Frames List:  
** `r names`  

## Subset Data  
Now that we have verified that the files were correctly read, we need to subset them based on the state of NC's FIPS code:  
```{r subsetState, echo=TRUE, warning=FALSE,error=FALSE, message=FALSE}
# Subset the df's where ST=='37' for fips code of State of North Carolina
for (i in 1:length(names)){
        b <- paste0(names[i], ' <-', names[i], '[', names[i], '$ST==\'37\',]')
        eval(parse( text=b ))
}
```

## Merge Data  
Next, we need to merge the catch and trip data frames by ID_CODE. This will create one data frame with corresponding trip and catch data. Once the dataframes are merged, we need to do a little housekeeping, by dropping unused factors fromt he dataset, and arranging the data by descending ID_CODE.  

``` {r mergedata, echo=TRUE, warning=FALSE,error=FALSE, message=FALSE}

# This function merges the df's passed to it based on the ID_CODE variable
mergeDf <- function(catch,trip, i){
        b <- paste0('merged.w',i, ' <- merge(', paste0(catch),',', paste0(trip), ', by=\'ID_CODE\')') 
        print(b)
        return(eval(parse( text=b )))
}

# create dataframes and populate them by passing to mergedf. Then arrange them by id_code. For each wave.
for (i in 1:6){
        d <- paste0("merged.w",i," <- data.frame()")
        a <- paste0("merged.w",i, " <- mergeDf(paste0('catch_', year, '",i,"'), paste0('trip_', year, '",i,"'), ",i,")")
        b <- paste0("merged.w",i,"$common <- factor(merged.w",i,"$common)")
        c <- paste0("merged.w",i, " <- arrange(merged.w",i,", desc(ID_CODE))")
        eval(parse( text=d ))
        eval(parse( text=a ))
        eval(parse( text=b ))
        eval(parse( text=c ))        
}
```


## Subset Data by County  
Now we need to subset the merged data by NC Counties.  
```{r subByCounty, echo=TRUE, warning=FALSE,error=FALSE, message=FALSE}
# create vectors of county codes to subset by
south <- c(19, 95, 129, 133, 141, 147)
north <- c(15, 29, 41, 53, 55, 139, 143, 177, 187)
central <- c(13, 31, 49, 137)

for (i in 1:6){
        #subset by county code into north and south NC
        a <- paste0("merged.w",i,".south <- merged.w",i,"[merged.w",i,"$CNTY %in% south,]")
        b <- paste0("merged.w",i,".north <- merged.w",i,"[merged.w",i,"$CNTY %in% north,]")
        c <- paste0("merged.w",i,".central <- merged.w",i,"[merged.w",i,"$CNTY %in% central,]")
        # Parse and eval all the strings
        eval(parse( text=a ))
        eval(parse( text=b ))
        eval(parse( text=c ))
        
        # Redefine the common names as factors. This will drop unused levels.
        d <- paste0("merged.w",i,".south$common <- factor(merged.w",i,".south$common)")
        e <- paste0("merged.w",i,".north$common <- factor(merged.w",i,".north$common)")
        f <- paste0("merged.w",i,".central$common <- factor(merged.w",i,".central$common)")
        eval(parse( text=d ))
        eval(parse( text=e ))
        eval(parse( text=f ))
        
        # create a vector of fish names caught from each area
        # I used m here since I couldnt reuse i
        g <- paste0("unique.w",i,".south <- levels(merged.w",i,".south$common)")
        h <- paste0("unique.w",i,".north <- levels(merged.w",i,".north$common)")
        m <- paste0("unique.w",i,".central <- levels(merged.w",i,".central$common)")
        eval(parse( text=g ))
        eval(parse( text=h ))
        eval(parse( text=m ))
        
        # Create a vector of the number of unique fish species caught
        j <- paste0("num.caught.south.w",i," <- length(unique.w",i,".south)")
        k <- paste0("num.caught.north.w",i," <- length(unique.w",i,".north)")
        l <- paste0("num.caught.central.w",i," <- length(unique.w",i,".central)")
        eval(parse( text=j ))
        eval(parse( text=k ))
        eval(parse( text=l ))
}
```

#Exploratory Analysis   
With the processing done, we can start to analyze the data.  

## Number of species reported per region  
This includes all catch: A+B1+B2  

```{r speciesDiversity, echo=TRUE, warning=FALSE,error=FALSE, message=FALSE}
library(reshape2)
library(ggplot2)

# Create regional DF's for catch
fish.count.central <- data.frame(num.caught.central.w1,num.caught.central.w2, 
                                 num.caught.central.w3, num.caught.central.w4, num.caught.central.w5)
names(fish.count.central) <- c("1", "2","3","4","5")
fish.count.central <- melt(fish.count.central)
names(fish.count.central) <- c("Wave","Species_Count")
# Create regional DF's for catch
fish.count.north <- data.frame(num.caught.north.w1,num.caught.north.w2, 
                               num.caught.north.w3, num.caught.north.w4, num.caught.north.w5)
names(fish.count.north) <- c("1", "2","3","4","5")
fish.count.north <- melt(fish.count.north)
names(fish.count.north) <- c("Wave","Species_Count")
# Create regional DF's for catch
fish.count.south <- data.frame(num.caught.south.w1,num.caught.south.w2, 
                               num.caught.south.w3, num.caught.south.w4, num.caught.south.w5)
names(fish.count.south) <- c("1", "2","3","4","5")
fish.count.south <- melt(fish.count.south)
names(fish.count.south) <- c("Wave","Species_Count")

# Merge the datasets into one master NC df
fish.count.1 <- merge(fish.count.north, fish.count.central, by="Wave")
fish.count <- merge(fish.count.1, fish.count.south, by="Wave")
names(fish.count) <- c("Wave", "North", "Central", "South")
# Melt dataset into longform, tidy, dataset
fish.count <- melt(fish.count)
# Rename the variables
names(fish.count) <- c("Wave", "Region","Count")

plot2 <- ggplot(fish.count, aes(Wave, Count, fill=Wave)) + 
        geom_bar(colour="black", stat="identity") +
        scale_fill_brewer() +
        facet_wrap(~ Region)+
        xlab("Wave") + ylab("Number of Species Caught")+
        ggtitle(paste("NC Species Diversity by Wave and Region", 
                      merged.w1.central$YEAR.x[1]))
print(plot2)
```

There does not appear to be much difference in the number of species identified by region. The Northern region had a mean of: `r mean(fish.count.north$Species_Count)`, while the south had a yearly mean of: `r mean(fish.count.south$Species_Count)`, and the central region has a yearly mean of: `r mean(fish.count.central$Species_Count)` Although small diffrence exist, these can be attributed to any number of factors other than interviewers skill at identifying fishes. The yearly mean species count for the state of NC in `r merged.w1.central$YEAR.x[1]` is: `r mean(fish.count$Count)`  
  
## Percentage of Fish Measured  

One useful metric of performance can be found in the ratio of fishes measured (Type A) vs harvest (A+B1). The concept is that the interviewer is accounting for all harvest, but not observing all of them. This could be for a number of confounding reasons, including angler cooperation, or weather; however, any region trending less than 50% over the course of the year should raise some questions.  

Here we show a snippet of code that we will run to compile this information. This snippet is for all waves across the Central region, and in the backgroud we will run all waves for all regions.  
```{r percMeasured, echo=TRUE, warning=FALSE,error=FALSE, message=FALSE }

for(i in 1:6){
        a <- paste0("a.w",i, "<- sum(merged.w",i,".central$HARVEST)")
        b <- paste0("b.w",i, "<- sum(merged.w",i,".central$CLAIM)")
        b2 <- paste0("b2.w",i, "<- sum(merged.w",i,".central$RELEASE)")
        eval(parse( text=a))
        eval(parse( text=b))
        eval(parse( text=b2))
}
```

Next we create a dataframe of all of the variables collected:  
```{r createDF, echo=TRUE, warning=FALSE,error=FALSE, message=FALSE}
catch <- c(a.w1,b.w1,b2.w1,                        
           a.w2,b.w2,b2.w2,
           a.w3,b.w3,b2.w3,
           a.w4,b.w4,b2.w4,
           a.w5,b.w5,b2.w5,
           a.w6,b.w6,b2.w6)

type <- c("A","B1","B2",
          "A","B1","B2",
          "A","B1","B2",
          "A","B1","B2",
          "A","B1","B2",
          "A","B1","B2")

location <- c("Central","Central","Central",
              "Central","Central","Central",
              "Central","Central","Central",
              "Central","Central","Central",
              "Central","Central","Central",
              "Central","Central","Central")

wave <- c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6)

# Add Wave1 to the df
catch.df <- data.frame(type,catch,location,wave)
```

```{r createDFNoShow, echo=FALSE, warning=FALSE,error=FALSE, message=FALSE}
for(i in 1:6){
        a <- paste0("a.w",i, "<- sum(merged.w",i,".south$HARVEST)")
        b <- paste0("b.w",i, "<- sum(merged.w",i,".south$CLAIM)")
        b2 <- paste0("b2.w",i, "<- sum(merged.w",i,".south$RELEASE)")
        eval(parse( text=a))
        eval(parse( text=b))
        eval(parse( text=b2))
}
catch <- c(a.w1,b.w1,b2.w1,                        
           a.w2,b.w2,b2.w2,
           a.w3,b.w3,b2.w3,
           a.w4,b.w4,b2.w4,
           a.w5,b.w5,b2.w5,
           a.w6,b.w6,b2.w6)

location <- c("South","South","South",
              "South","South","South",
              "South","South","South",
              "South","South","South",
              "South","South","South",
              "South","South","South")

# tempdf to hold the south data 
temp.df <- data.frame(type,catch,location,wave)
# row bind the central and southern data
catch.df <- rbind(catch.df,temp.df)

for(i in 1:6){
        a <- paste0("a.w",i, "<- sum(merged.w",i,".north$HARVEST)")
        b <- paste0("b.w",i, "<- sum(merged.w",i,".north$CLAIM)")
        b2 <- paste0("b2.w",i, "<- sum(merged.w",i,".north$RELEASE)")
        eval(parse( text=a))
        eval(parse( text=b))
        eval(parse( text=b2))
}
catch <- c(a.w1,b.w1,b2.w1,                        
           a.w2,b.w2,b2.w2,
           a.w3,b.w3,b2.w3,
           a.w4,b.w4,b2.w4,
           a.w5,b.w5,b2.w5,
           a.w6,b.w6,b2.w6)

location <- c("North","North","North",
              "North","North","North",
              "North","North","North",
              "North","North","North",
              "North","North","North",
              "North","North","North")


temp.df <- data.frame(type,catch,location,wave)

catch.df <- rbind(catch.df,temp.df)
names(catch.df) <- c("Type","Count","Region","Wave")
```
Lastly, we will plot the catch type by region:  
```{r percPlot, echo=TRUE, warning=FALSE,error=FALSE, message=FALSE}
library(ggplot2)
plot2 <- ggplot(catch.df, aes(Wave, Count, fill=Type)) + 
        geom_bar(colour="black", stat="identity") +
        #scale_fill_brewer() +
        facet_wrap(~ Region)+
        xlab("Wave") + ylab("Number of Fishes Caught")+
        ggtitle(paste("NC-MRIP Type A,B1,B2 Catch \nby Wave and Region", 
                      merged.w1.central$YEAR.x[1]))
print(plot2)
```

To further explore the differences in fish landed vs fish landed and measured, we present the data as the percent of total catch measured by region.  

First off, we create a function to return the catch values for each type, A, B1, B2:  

```{r percTotCatchMeasured, echo=TRUE, warning=FALSE,error=FALSE, message=FALSE}
calcPercent <- function(wave, type, region){
        A <- (catch.df[catch.df$Wave==wave & catch.df$Type==type & catch.df$Region==region,2])
        return(A)
}
```

Then we feed the catch data from each region and wave to the function. Displayed here are just the data from the central region, the rest of the data will be crunched in the background:  

```{r perTotCatchCont, echo=TRUE, warning=FALSE,error=FALSE, message=FALSE}
for (i in 1:6){
        A <- calcPercent(i,"A","Central")
        B1 <- calcPercent(i,"B1","Central")
        B2 <- calcPercent(i,"B2","Central")
        b <- paste0("percent.central.w",i, "<- (A/(A+B1)*100)")
        eval(parse( text=b))
}
```

```{r perTotCatchContHidden, echo=FALSE, warning=FALSE,error=FALSE, message=FALSE}
for (i in 1:6){
        A <- calcPercent(i,"A","South")
        B1 <- calcPercent(i,"B1","South")
        B2 <- calcPercent(i,"B2","South")
        b <- paste0("percent.south.w",i, "<- (A/(A+B1)*100)")
        eval(parse( text=b))
}

for (i in 1:6){
        A <- calcPercent(i,"A","North")
        B1 <- calcPercent(i,"B1","North")
        B2 <- calcPercent(i,"B2","North")
        b <- paste0("percent.north.w",i, "<- (A/(A+B1)*100)")
        eval(parse( text=b))
}
```

Lastly, we create and melt a dataframe into a plottable format:  
```{r cleanCatchPercPer, echo=TRUE, warning=FALSE,error=FALSE, message=FALSE}
# Create vectors of the calculated percentages for each region
percentscentral <- c(percent.central.w1,percent.central.w2,percent.central.w3,percent.central.w4,percent.central.w5,percent.central.w6)
percentssouth <- c(percent.south.w1,percent.south.w2,percent.south.w3,percent.south.w4,percent.south.w5,percent.south.w6)
percentsnorth <- c(percent.north.w1,percent.north.w2,percent.north.w3,percent.north.w4,percent.north.w5,percent.north.w6)
# Include the corresponding waves
waves <- c(1,2,3,4,5,6)

# Create a dataframe from the vectors
percents.tot <- data.frame(waves, percentscentral, percentssouth, percentsnorth)
names(percents.tot) <- c("Wave", "Central","South","North")

# Rearrange (melt) the dataframe into long format. This is prefered by ggplot.
percents.tot <- melt(percents.tot, id="Wave")
names(percents.tot) <- c("Wave", "Region","Percent")
```

Then we plot the data:  
```{r plotCatchPercPer, echo=TRUE, warning=FALSE,error=FALSE, message=FALSE}
plot3 <- ggplot(percents.tot, aes(Wave, Percent, fill=Wave)) + 
        geom_bar(colour="black", stat="identity") +
        #scale_fill_brewer() +
        facet_wrap(~ Region)+
        xlab("Wave") + ylab("Percent of Total Catch Measured(A/(A+B1))*100")+
        ggtitle(paste("NC-MRIP Percent of Total Catch Measured \nby Wave and Region", merged.w1.central$YEAR.x[1]))
print(plot3)
```

##Number of Intercepts by Mode and Region  
In this section we show the number of intercepts acheived per mode by region.  

